---
title: "P&S-2025: Lab assignment 4"
author: "Yustyna Hayevska, Oleksii Lasiichuk, Ivan Zarytskyi"
output:
  html_document:
    df_print: paged
---

### Work breakdown:

-   Yustyna Hayevska (Task 3)

-   Oleksii Lasiichuk (Task 1, 2)

-   Ivan Zarytskyi (Task 4)

### Data generation
```{r}
# team ID
n <- 13

# function gets fractional part of a number (5.7 -> 0.7)
get_fractional <- function(x) {
  return(x - floor(x))
}

# 100 values of x, and 50 of y (150 total)
k_values <- 1:150

# apply formula from the assignment
raw_values <- k_values * log(k_values^2 * n + pi)

a_values <- get_fractional(raw_values)

X <- qnorm(a_values[1:100])
Y <- qnorm(a_values[101:150])
```


### Task 1
Hypothesis:
$$H_0: \mu_1 = \mu_2$$
$$H_1: \mu_1 \neq \mu_2$$

$\sigma_1^2 = \sigma_2^2 = 1$ (Variances are known).

We are testing the equality of means between two independent populations. Since the population variances are known ($\sigma^2 = 1$), the two sample Z-test is the appropriate standard test. (If variances were unknown, we would use a T-test).

For a significance level $\alpha = 0.05$, we reject $H_0$ if $|Z_{stat}| > Z_{0.025}$.

$Z_{0.025} \approx 1.96$.
So rejection region is: $(-\infty, -1.96) \cup (1.96, \infty)$.

```{r}
sigma1_sq <- 1
sigma2_sq <- 1
n_x <- 100
n_y <- 50
alpha <- 0.05

mean_x <- mean(X)
mean_y <- mean(Y)

# Z-test statistic calculation
# formula: z = (mean_x - mean_y) / sqrt(sigma1^2/n_x + sigma2^2/n_y)
standard_error <- sqrt((sigma1_sq / n_x) + (sigma2_sq / n_y))
z_stat <- (mean_x - mean_y) / standard_error

# P-value for two-sided test: p = 2 * P(Z > |z_stat|)
p_value_z <- 2 * (1 - pnorm(abs(z_stat)))


cat("Sample mean X:", mean_x, "\n")
cat("Sample mean Y:", mean_y, "\n")
cat("Z-Statistic:", z_stat, "\n")
cat("P-value:", p_value_z, "\n")

```

Since the p-value ($0.663$) is significantly higher than our chosen significance level ($\alpha = 0.05$), we fail to reject the null hypothesis.

This indicates that the difference between the sample means ($0.0198$ vs $-0.0557$) is statistically insignificant. There is no evidence that populations for $X$ and $Y$ have different means. 

### Task 2
Hypothesis:
$$H_0: \sigma_1^2 = \sigma_2^2$$
$$H_1: \sigma_1^2 > \sigma_2^2$$
$\mu_1$ and $\mu_2$ (means are unknown)

We are going to use F-test for equality of variances. When testing the equality of variances from two normal populations with unknown means, the ratio of sample variances follows an F-distribution.

Rejection Region:This is a one-sided test (greater than). We reject $H_0$ if $F_{stat} > F_{\alpha, n_x-1, n_y-1}$.For $\alpha=0.05$, $df_1 = 99$, $df_2 = 49$.

```{r}
var_x <- var(X)
var_y <- var(Y)

#F-Test formula: F = S_x^2 / S_y^2
f_stat <- var_x / var_y

# P-value

# One-sided test (H1: sigma1^2 > sigma2^2): P(F > f_stat)
# In R, pf() gives P(F < q), so we use 1 - pf()
p_value_f <- 1 - pf(f_stat, n_x - 1, n_y - 1)

# Critical value for rejection region
f_critical <- qf(1 - alpha, n_x - 1, n_y - 1)

cat("Sample Variance X:", var_x, "\n")
cat("Sample Variance Y:", var_y, "\n")
cat("F-Statistic:", f_stat, "\n")
cat("Critical Value:", f_critical, "\n")
cat("P-value:", p_value_f, "\n")

```

We conducted a one-sided F-test to check if the variance of population X is greater than population Y ($H_0: \sigma_1^2 = \sigma_2^2$ vs $H_1: \sigma_1^2 > \sigma_2^2$).

The calculated F-statistic is 1.367, which is less than the critical value of 1.531. Correspondingly, the p-value is 0.113

Since the p-value ($0.113$) is higher than our chosen significance level ($\alpha = 0.05$), we fail to reject the null hypothesis.This indicates that while the sample variance of X ($1.10$) appears larger than Y ($0.80$), this difference is statistically insignificant. The data does not provide sufficient evidence to claim that the true population variance of X is greater than that of Y.

### Task 3
...

### Task 4
...
